Validation of Visual Statistical Inference, Applied to Linear Models

### 2012-06-09 paper_jasa.tex
lemma 3.1 final? - figure 3 fixed

The probability of picking the data plot from a lineup depends on the size $m$ of the lineup and the strength of the signal in the data plot. 
Under the above assumptions, we can express the probability as:
\[
P(p_D < p_0) = 1 - E\left[ (1 - p_D)^{m-1}\right]
\]
where $p_D$ is the $p$-value associated with the data in the test statistic, and $p_0$ is the minimum of all $p$-values in the data going into null plots.
\end{lemma}

@@

\hh{Their minimum $p_0 = \min_{1 \le i \le m-1}  \ p_{0,i}$ then follows a Beta distribution with shape parameters 1 and $m-1$, and corresponding distribution function 
\[
F_{p_0} (x) =  (1- x) ^{m-1} \text{ for } x \in [0,1].
\]

\begin{eqnarray*}
P(p_D < p_{0}) &=& 1 - P(p_{0} \le p_D) = 1- \int_0^1  P(p_{0} \le p_D \mid p_D=t) f_{p_D}(t) dt =  \\
&=& 1 - \int_0^1 F_{p_{0}}(t) f_{p_D}(t) dt = 1 - \int_0^1 (1-t)^{m-1} f_{p_D}(t) dt = 1 - E\left[ (1 - p_D)^{m-1}\right],
\end{eqnarray*}

}
%
%
%
%Let us think of a lineup as a head-to-head comparison of the test statistic and $m-1$ null plots. 
%%Let the data plot have a $p$-value of $p_D$, and the null plots $p$ values of $p_{0, i}$ with $1 \le i \le m-1$.
%We know that for each comparison the probability that the data, on which the plot is based, has the smaller $p$ value is 
%\begin{eqnarray*}
%P(p_D < p_{0,i}) &=& 1 - P(p_{i,0} \le p_D) = 1- \int_0^1  P(p_{i,0} \le p_D \mid p_D=t) f_{p_D}(t) dt =  \\
%&=& 1 - \int_0^1 F_{p_{0,i}}(t) f_{p_D}(t) dt = 1 - \int_0^1 t f_{p_D}(t) dt = 1 - E[p_D],
%\end{eqnarray*}
%which, in particular, is independent of $p_{0,i}$ for all $i$.
%
%Let us make the assumption that an observer is able to identify  the chart corresponding to the data with the smallest $p$-value. Further we will assume that all observers  have the same ability in identifying the data plot.
%
%With that, we define $Z$  as the number of null plots in a lineup, for which the $p$-value $p_{0,i}$ is smaller than $p_D$.  
%
%Then $Z \sim B_{m-1, E[p_D]}$, and the probability that an observer will pick the data plot in a given lineup is 
%\[
%P(Z=0) = \left(1 - E[p_D] \right)^{m-1} = P(p_D \le p_0), \ \ \ \text{ where } p_0 = \min_{1 \le i \le m-1}  \ p_{0,i}.
%\]
\end{proof}

@@

Now assume that the same lineup plot is evaluated by $K$ independent observers. W.l.o.g.\green{Can we get rid of cliches like WLOG?} let the last plot be the data plot and plots 1 through $m-1$ the null plots. Let $n_i$ be the number of observers who picked plot $i$ as the plot with the structure least consistent with the null hypothesis, corresponding to random variable $N_i$.  Then  $(N_1, N_2, ..., N_m) \sim$ Mult$_{\pi_1, ..., \pi_m}$ with $\sum_i \pi_i = 1$ and $\sum_{k=1}^{m} n_k = K$. We can estimate $\pi_i$ as $\widehat{\pi_i} = n_i/K$.

For the distribution of $N_m$, %the number of times out of $K$ that observers picked the last plot to show the structure least consistent with the null hypothesis, 
we get
\[
N_m\sim \left \{ 
\begin{array}{ll}
B_{K, (1-E[p_D])^{m-1}} & \text { under } H_1\\
B_{K, 1/m} & \text { under } H_0
\end{array}
\right .
\]

By matching the expected values, we therefore get an expression for the $p$-value in the data under $H_1$  as:
\begin{equation}\label{eqn:power_estimate}
{E[p_D]} = 1- \left(\frac{E[N_m]}{K}\right)^{1/(m-1)}.
\end{equation}

Note that the right hand side in Equation \ref{eqn:power_estimate} is independent of the parameter $\theta$, but just based on the lineup evaluation by independent observers. While this allows us to compute this value independently of the parametric setting, it also means that observers might pick out the data plot for a reason not related to the value of $\theta$.

>>>>

The probability of picking the data plot from a lineup depends on the size $m$ of the lineup and the strength of the signal in the data plot. 
Under the above assumptions, we can express the probability as:
\[
P(p_D < p_0) =  E\left[ (1 - p_D)^{m-1}\right]
\]
where $p_D$ is the $p$-value associated with the data in the test statistic, and $p_0$ is the minimum of all $p$-values in the data going into null plots.
\end{lemma}

@@

\hh{Their minimum $p_0 = \min_{1 \le i \le m-1}  \ p_{0,i}$ then follows a Beta distribution with shape parameters 1 and $m-1$, and corresponding distribution function 
\[
F_{p_0} (x) =  1- (1- x) ^{m-1}  \text{ for } x \in [0,1].
\]

\begin{eqnarray*}
P(p_D < p_{0}) &=& 1 - P(p_{0} \le p_D) = 1- \int_0^1  P(p_{0} \le p_D \mid p_D=t) f_{p_D}(t) dt =  \\
&=& 1 - \int_0^1 F_{p_{0}}(t) f_{p_D}(t) dt = 1 - \int_0^1 f_{p_D}(t) dt + \int_0^1 (1-t)^{m-1} f_{p_D}(t) dt =  E\left[ (1 - p_D)^{m-1}\right].
\end{eqnarray*}

This proves the statement above. We further see:
\[
E\left[ (1 - p_D)^{m-1}\right] =  \sum_{k=0}^{m-1} {m-1 \choose k} (-E[p_D])^{k} = 1-(m-1)E[p_D] + O(E[p_D^2]).
\]
This implies that as $p_D$, the $p$-value, becomes smaller, the probability that the observer identifies the data plot goes up and,
conversely, as $m$, the number of choices given in the lineup, goes up the probability to pick the data plot  goes down. 
}
%
%
%
%Let us think of a lineup as a head-to-head comparison of the test statistic and $m-1$ null plots. 
%%Let the data plot have a $p$-value of $p_D$, and the null plots $p$ values of $p_{0, i}$ with $1 \le i \le m-1$.
%We know that for each comparison the probability that the data, on which the plot is based, has the smaller $p$ value is 
%\begin{eqnarray*}
%P(p_D < p_{0,i}) &=& 1 - P(p_{i,0} \le p_D) = 1- \int_0^1  P(p_{i,0} \le p_D \mid p_D=t) f_{p_D}(t) dt =  \\
%&=& 1 - \int_0^1 F_{p_{0,i}}(t) f_{p_D}(t) dt = 1 - \int_0^1 t f_{p_D}(t) dt = 1 - E[p_D],
%\end{eqnarray*}
%which, in particular, is independent of $p_{0,i}$ for all $i$.
%
%Let us make the assumption that an observer is able to identify  the chart corresponding to the data with the smallest $p$-value. Further we will assume that all observers  have the same ability in identifying the data plot.
%
%With that, we define $Z$  as the number of null plots in a lineup, for which the $p$-value $p_{0,i}$ is smaller than $p_D$.  
%
%Then $Z \sim B_{m-1, E[p_D]}$, and the probability that an observer will pick the data plot in a given lineup is 
%\[
%P(Z=0) = \left(1 - E[p_D] \right)^{m-1} = P(p_D \le p_0), \ \ \ \text{ where } p_0 = \min_{1 \le i \le m-1}  \ p_{0,i}.
%\]
\end{proof}

@@

Now assume that the same lineup plot is evaluated by $K$ independent observers. %W.l.o.g.\green{Can we get rid of cliches like WLOG?} 
Let the last plot be the data plot and plots 1 through $m-1$ the null plots. Let $n_i$ be the number of observers who picked plot $i$ as the plot with the structure least consistent with the null hypothesis, corresponding to random variable $N_i$.  Then  $(N_1, N_2, ..., N_m) \sim$ Mult$_{\pi_1, ..., \pi_m}$ with $\sum_i \pi_i = 1$ and $\sum_{k=1}^{m} n_k = K$. We can estimate $\pi_i$ as $\widehat{\pi_i} = n_i/K$.

For the distribution of $N_m$, %the number of times out of $K$ that observers picked the last plot to show the structure least consistent with the null hypothesis, 
we get
\[
N_m\sim \left \{ 
\begin{array}{ll}
B_{K, \hh{E\left[ (1 - p_D)^{m-1}\right]}} & \text { under } H_1\\
B_{K, 1/m} & \text { under } H_0
\end{array}
\right .
\]

By matching the expected values, we therefore get an expression for the $p$-value in the data under $H_1$  as:
\begin{equation}\label{eqn:power_estimate}
%{E[p_D]} = 1- \left(\frac{E[N_m]}{K}\right)^{1/(m-1)}.
\hh{1 - E[N_m]/K = (m-1)E[p_D] + O(E[p_D^2])}
\end{equation}

Note that the \hh{left}  hand side in Equation \ref{eqn:power_estimate} is independent of the parameter $\theta$, but just based on the lineup evaluation by independent observers. While this allows us to compute this value independently of the parametric setting, it also means that observers might pick out the data plot for a reason not related to the value of $\theta$.
